---
title: "6315305_assignment_eda"
author: "Andreas Sakapetis"
date: "November 30, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r}
#loading the library "readxl" to be able to use the function read_xlsx().
library(readxl)
#loading the library "formattable" to introduce a table with all the variables.
library(formattable)
#loading the library "ggplot2" to create usefull plots.
library(ggplot2)
#loading the library "GGally" to create usefull plots.
library(GGally)
#loading the library "tidyverse" to perform varius manipulations.
library(tidyverse)

```
```{r}
#importing data from local folder to Rstudio in ob.data variable.
ob.data <- read_xlsx("C:\\Users\\Andreas\\Desktop\\Assignment A\\PROBESITY dataset.xlsx", sheet = 1)
#import the first sheet with the variable abbreviations.
var_explain = read_xlsx("C:\\Users\\Andreas\\Desktop\\Assignment A\\PROBESITY dataset.xlsx", sheet = 2) 
```

```{r}
#view the names of the columns in the ob.data dataset.
names(ob.data)
```


```{r}
#Use the function head() to examine at the first few rows of the ob.data dataset.
head(ob.data)

```

```{r}
#Use the function tail() to examine the last few rows of the ob.data dataset.
tail(ob.data)
```

The dataset consists of 19 variables, with a total of 1890 observations.The data 
includes information about prevalence rates for female and male adults 
(aged 20 and over) from 45 countries in the 1975-2016 period. There are 3054 
missing values, spread across different variables. Within the datase exist variables
that the type is either numeric or character. However, the variable euro45 and eu
are categorical variables. More specific for the variable euro45 if the value is
1 this means that the country is a member of the european union, else if the value 
is 0 then the country is not. As for the eu variable if the the value is 1, this 
is translated that the country is a region of the European continent, else if the
value is 0 the country is outised the European continent. Moreover, within the 
dataset are variables that are calculated in order to assist the researchers of 
the study to make conlusions such as urban, demeaned_lhc, demeaned_urban, 
irgdppop_lhp and loga.


```{r}
#This function counts missing values
sapply(ob.data, function(x) sum(length(which(is.na(x)))))
```


```{r}

#create a table with all the variables 
widget.formattable = formattable(var_explain) 
#display the table 
widget.formattable 

```

```{r}
#With sapply function it is possible to examine the type of the variables.
sapply(ob.data,class)
```

```{r}
#acquire summary statistics for the ob.data dataset.
summary(ob.data)
```



```{r}
hist(ob.data$hc)
```
The histogram for Human capital index, based on years of schooling and returns 
to education reveals that there is a negative skewness.

```{r}

hist(ob.data$urban)

```

```{r}
#With this lines the dataset is splitted in two for better visualization 
#purposes. First, making sure that the dataset is sorted alphabetically by the
#country variable. Then the dataset is splitted in two halves. Manual inspection
#assisted the effort to distinguish the precise row that the dataset needs to be 
#splitted.

#Alphabetically sorting the dataset based on the country variable.
ob.data <- ob.data %>%
  arrange(ob.data$country)

#This is the first half.
ob.data.half1 <-ob.data[1:966,]

#This is the second half.
ob.data.half2 <-ob.data[967:1890,]
```



```{r}
ob.data.half1 %>% 
  ggplot(aes(x = ob.data.half1$country, y = ob.data.half1$PFOV20P, fill = ob.data.half1$country)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()


```

```{r}
ob.data.half2 %>% 
  ggplot(aes(x = ob.data.half2$country, y = ob.data.half2$PFOV20P, fill = ob.data.half2$country)) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
```







```{r}
#creating a more refined dataframe with the variables that seem to be more 
#interesting.
euro.obese <- 
  ob.data %>%
  select(year, code, eu, PFOV20P, PFOB20P, PMOV20P, PMOB20P) %>%
  mutate(obesity.sum = PFOB20P + PMOB20P)

#Transforming the eu variable to factor.
euro.obese$eu <- as.factor(euro.obese$eu)

#Creating a plot to examine the frequency of obesity in the Euopean Continent and
#outside the European Continent.
frequency.obese <- 
  euro.obese %>%
  ggplot(aes(x = obesity.sum, linetype  = eu))+
  geom_freqpoly(size = 1, bins = 60 , color = "blue")+
  theme_minimal()

frequency.obese 
```

```{r}
#Creating a plot to examine how obesity transforms into the given timeline
#separatedi n the Euopean Continent and outside the European Continent.
total.obesity <- 
  euro.obese  %>%
  ggplot(mapping = aes(x = year, y = obesity.sum, linetype = eu))+
  geom_smooth()+
  theme_minimal()

total.obesity
```

```{r}
#Creating a plot to examine how female obesity transforms into the given timeline
#separatedi n the Euopean Continent and outside the European Continent.
female.obesity <- 
  euro.obese %>%
  ggplot(mapping = aes(x = year, y = PFOB20P, linetype = eu))+
  geom_smooth()+
  theme_minimal()

female.obesity

```

```{r}
#Creating a plot to examine how male obesity transforms into the given timeline
#separatedi n the Euopean Continent and outside the European Continent.
male.obesity <- 
  euro.obese  %>%
  ggplot(mapping = aes(x = year, y = PMOB20P, linetype = eu))+
  geom_smooth()+
  theme_minimal()

male.obesity
```

```{r}
#Creating a plot to compare Female and Male obesity.
hist(euro.obese$PFOB20P, col=rgb(1,0,0,0.5), main= "Overlapping Histogram", xlab="Variable")
hist(euro.obese$PMOB20P, col=rgb(0,0,1,0.5), add=T)
box()
legend("topright", legend=c("Female Obesity", "Male Obesity"),
       col=c("red", "blue"), lwd=4, cex=0.8)
```


```{r}
#Creating a plot to compare Females and Males that are overweight. 
hist(euro.obese$PFOV20P, col=rgb(1,0,0,0.5), main= "Overlapping Histogram", xlab="Variable")
hist(euro.obese$PMOV20P, col=rgb(0,0,1,0.5), add=T)
box()
legend("topright", legend=c("Female Obesity", "Male Obesity"),
       col=c("red", "blue"), lwd=4, cex=0.8)
```

From the careful obseravation of varius graphics it is possible to conclude that
the total population that is obese increases throughout the years. Taking into
consideration the given data, the population in countries of the European Continent
tends to be more obese than the population in the countries outside the European 
Continent. Another important fact is that obesity is more prone to females than 
males.


```{r}
#Summarizing the countries based on obesity.sum.
obese.country <- 
  euro.obese %>%
  group_by(code) %>%
  summarize(
    sum(obesity.sum)
  ) 

colnames(obese.country) <- c("Country", "Total_Obesity")

#Create a plot to examine which are the most obese countries.
obesity.plot <-
  obese.country %>%
  ggplot(mapping = aes(x = Country, y = Total_Obesity, fill= Country))+
  geom_bar(stat = "identity", width = 0.3) +
  theme(axis.text.y = element_text(size = rel(0.9), angle = 0))+
  coord_flip() 

obesity.plot

```

  As it possible to Observe from the graphic there are countries of the European 
  continent that surpass in obesity countries from outside the European continent.
  This fact comes to a total agreement with the conclusions made so far from the
  previous graphics.
  
  
Firstly, the analyst loaded the necessary libraries readxl, formattable, ggplot2,
ggally, tidyverse each designated for a different purpose. Then the analyst 
proceeds with loading the data from a local file location. In order to get a firm,
grasp of the data the analyst used a series of functions such as: names, 
head and tail. 

After the analyst used an sapply, a sum, a length, a which and an 
is.na function in conjunction to find the missing values of the dataset. 
The function formattable was utilized to load the second sheet of excel file 
which contained information about the variables of the dataset. Then With sapply 
function it is possible to examine the type of the variables. 

Summary statistics  were calculated by summary function. Moreover, various 
histograms were utilized  to see the tendencies of the variables and their 
distribution. For better visualization purposes the analyst divided the dataset 
in two halves based on alphabetic order. 

Thus, the boxplots were created to be able to examine the  countries based on 
obese population, no outliers were spotted. Furthermore, a 
frequency diagram was created to analyze the frequency of obese people in and 
outside the European Continent. 

Apart from that. The created line diagrams  depict how obesity is related to 
the timeline provided by the dataset. The overlapping histograms were created to
show the difference between men and women independent of the countries. 
Finally the bar plot depicts the total obesity for the countriescalculated as 
the total sum of women and men.

